---
title: "National Air Quality and Its Effects on Adult Asthma, A Time Series Analysis"
name: "Rebecca Madsen"
class: "DATA 470"
output: html_notebook
---

## Abstract

Air quality and asthma, among other respiratory illnesses, are a continued struggle for many communities. In addition, the continued growth of the national and global population warrants our nation's pollution to continue to increase. This, in turn, will give to a worsening in respiratory conditions and an increase of respiratory troubles, such as asthma. Using data provided by the CDC National Environmental Public Health Tracking Network Query Tool, statistical analysis was used to see if there is a relationship between the average number of days above PM2.5 standard levels nationally and the percent of adults diagnosed with asthma who report they currently have asthma nationally through the period of 2011 to 2016. The variables were looked at both individually and against each other with the period established using the statistical tool R. Basic statistical analysis and time series modeling were used as a way of testing for a relationship between air quality and asthma. 

### Introduction

Respiratory Illnesses are greatly affected by high pollution, making it one of the greatest problems that the modern world faces. (CDC, 2017) Asthma being one of the biggest and more commonly known respiratory illness is greatly affected by the air quality. Particles that have a diameter that measures 2.5 micrometers, and remain suspended longer in the air which causes pollution. These are particles that come from the burning of fuel and chemical reactions, worsening the air quality around us. (Airveda, 2017) Airveda explains that "exposure to PM2.5 has multiple short term and long term health impacts. Short term include irritation in the eyes, nose and throat, coughing, sneezing and shortness of breath. A prolonged exposure to PM2.5 can cause permanent respiratory problems such as asthma, chronic bronchitis and heart disease." (Airveda, 2017) The CDC is using its own national tracking network to follow many environmental and public health risks. This tracking tool includes information on both asthma and PM2.5 particulates at a national level. 

Using the CDC tracking network, I downloaded data on the percent of adults diagnosed with asthma nationally and the average days of PM2.5 levels above the standard nationally. My goal of this time series analysis is to explore if there is any correlation between numerical percentage of national average PM2.5 levels and the numerical percentage of national adults currently reporting asthma between the years 2011 to 2016. 

```{r}
library(mosaic)
library(broom)
library(tidyverse)
library(openintro)
library(Stat2Data)
library(car)


library(readr)
asthma_pmlevel <- read_csv("asthmaAndPMLevelCombined.csv")
View(asthma_pmlevel)
```


Since I am using time series analysis, I will need to upload and run a few packages to make this analysis possible. CRAN gives us a huge set of public packages, and happens to have some great packages that can be used in time series analysis. Since R already has basic time series installed because of CRAN, we will be using those to explore the data between the years of 2011 to 2016. 

### Methods

Below I am using some basic R codes to explore the dataset.

```{r}
str(asthma_pmlevel)
```


Above is general structure information about the data I am going to explore for my time series analysis of adult asthma and air quality. Next, I ran the summary statistics of the dataset for each variable (as seen below). However, the three variables I want to focus on are `Year`, `Asthma`, and `AvgPM2.5` for this analysis. 

```{r}
summary(asthma_pmlevel)
```

```{r}
ggplot(asthma_pmlevel, aes(y = Asthma, x = AvgPM2.5)) +
    geom_point()
```

Above, we can see there is right-skewedness from both the variables `AvgPM2.5` and `Asthma`. Filtering the variable `AvgPM2.5` variable below, to eliminate the outliers beyond 0.008.

```{r}
asthma_pmtrans <- filter(asthma_pmlevel, AvgPM2.5 > 0.0 & AvgPM2.5 < 0.08)
asthma_pmtrans
```

```{r}
ggplot(asthma_pmtrans, aes(y = Asthma, x = AvgPM2.5)) +
    geom_point() +
    geom_smooth(method = "lm")
```

I filtered the variable for `AvgPM2.5` which helped with some of the skewedness from outliers beyond 0.15 but there still seems to be some right-skew in the dataset. Next I am filtering the variable `Asthma` to eliminate outliers from before 0.07.

```{r}
asthmatrans_pmfilter <- filter(asthma_pmtrans, Asthma > 0.07)
asthmatrans_pmfilter
```

```{r}
ggplot(asthmatrans_pmfilter, aes(y = Asthma, x = AvgPM2.5)) +
    geom_point() +
    geom_smooth(method = "lm")
```
It helped a little, showing a slight negative trend. However, it is not enough to say there is a negative relationship between the two variables. 

Next, I transformed the variable `AvgPM2.5`, using the log function since it appears to be rightly skewed. I then used the tidy function to get the values and plotted the transformed data on a scatter plot along with a linear model fit. 

```{r}
asthma_pm_lm <- lm(log(AvgPM2.5) ~ Asthma, data = asthmatrans_pmfilter)
asthma_pm_lm_tidy <- tidy(asthma_pm_lm, conf.int = TRUE)
asthma_pm_lm_tidy
```

```{r}
#LAsthma <- log(asthma_pmlevel$Asthma)
LPM <- log(asthma_pmlevel$AvgPM2.5)

ggplot(asthma_pmlevel, aes(y = Asthma, x = LPM)) +
  geom_point() +
    geom_smooth(method = "lm")
```

This still gave us no association between the transformed variables of `AvgPM2.5` and the variable `Asthma`. The intercept p-value of 3.106e-07 tells us there is some significance when the intercept of `AvgPM2.5` is compared to `Asthma` unlike `Asthma` by itself which gives a p-value of 7.81e-01.

#### Beginning Time Series Analysis

First, I want to see if there is any association between the variables `Asthma` and `AvgPM2.5` separately with time between 2011 to 2016. 

```{r}
adult_asthma <- ts(asthmatrans_pmfilter$Asthma, start= 2011, end= 2016, frequency= 1)
plot(adult_asthma)

average_pm <- ts(asthmatrans_pmfilter$AvgPM2.5, start = 2011, end = 2016, frequency = 1)
plot(average_pm)
```
Above we are looking at two trend lines: the percent of adults with asthma and the average days of PM2.5 levels above standard separately against time. 

* Starting with the `adult_asthma` plot, we see a positive increase from 2011 to 2012, a plateau from 2012 to 2013, then a sharp decrease from 2013 to 2014, then a smaller increase between 2014 and 2015 before plateauing from 2015 to 2016. Across time, this tells us the numerical percentage of adults with currently reported slightly decrease from 2011 to 2016.

* Next, looking at `average_pm` we see a plateau in numerical percentage of days above national levels for PM2.5 particles from 2011 to 2013, then a increase from 2013 to 2016. Across time, this tells us that there was an increase in number of average days above PM2.5 standards from 2011 to 2016. 

Below I am using the `HoltWinters()`, a function to fit an exponential model for limiting the squared prediction error in time series forecasting, to the variables `average_pm` and `adult_asthma` separately. The red line is the exponential fit squared predicted error for the time series where the black line is the exponential fit for each time series.

```{r}
fit_pm <- HoltWinters(average_pm, gamma=FALSE)
plot(fit_pm)

fit_asthma <- HoltWinters(adult_asthma, gamma = FALSE)
plot(fit_asthma)
```

Using time series analysis, we also should look at creating lag. A lag of time series is created when any given number of periods in time are shifted. This is because the state of a period back in the time-line could still be affecting the current state of the time series. This will allow us to see if an earlier year is affecting a later year in our time series.

Below, I am looking at an autocorrelation for both `average_pm` and `adult_asthma` separately. The autocorrelation is looking at lag in itself for each variable to check for any correlation. The `average_pm` time series doesn't fall to 0 quickly telling us it is stationary with time. However, `adult_asthma` does fall fairly quickly to 0 in lag, telling us it is not stationary with time. Since both time series cross the positive blue dashed line, there is a correlation for that specific lag with the current series.

```{r}
acfRes_pm <- acf(average_pm)
acfRes_asthma <- acf(adult_asthma)
```

Lastly, I am looking at cross-correlation between both `adult_asthma` and `average_pm` series at once, to see if there is any causal relationship between the two time series variables. With cross-correlation, the closer the value is to 1 (either -1.0 or +0.1) the closer the sets are to being identical.

```{r}
ccfRes <- ccf(adult_asthma, average_pm, ylab = "cross-correlation")
```

### Results

As we can see above, the lags for the cross-correlation are not very close to 1 but to 0, telling us that the two time series are not identical. The points of lag also do not cross the dashed blue line, telling us there is no correlation between the two time series. 


### Discussion/Conclusion

Through a time series analysis, I found that there is no relationship between the percent of adults diagnosed with asthma stating it is still present and the percent of days above the standard PM2.5 levels through the years of 2011 to 2016. 

Beginning with the Holts-Winter exponential fits, we see a forecasting for each individual variable. The `fit_pm` graph shows overall positive association with the time period 2011-2016. The black line shows the exponential fit and the red line represents the standard predictor error, used for forecasting. From this, one can say there is the possibility of an increase in days above standard PM2.5 particulates nationally. The `fit_asthma` graph shows a more decrease and plateau for the exponential fit black line and a sharp decrease then slight increase of the standard predictor error shown as the red line. From this, one can say that the number of adults diagnosed with asthma that still currently report it nationally has neither an increase nor decrease. 

Next, the first spikes on the separate auto-correlations of the time-series variables `Asthma` and `PM2.5` shows a significant time lag for that year. (2011 starting at 0 going to 2016 at 5) The other spikes fall closer to zero, giving no significant time lag for the years 2012-2016. However, this only explains the two variables separately against a time correlation.

The cross-correlation `ccfRes` graph above tells us there is no association between `Asthma` and `PM2.5` because of the lags (spikes) not crossing the dashed blue lines. If the lags had crossed the dashed lines, we could see more significant association between the two time-series variables. One can also see that there is no correlation from each of the lags being close to zero, telling us there is no correlation between the two time-series variables for the years 2001 to 2016. 

Further statistical and time series analysis would be needed to explore a deeper look at the variables used, with a more in-depth time series and possible higher level analysis with a larger period if available.



#### Bibliography/Citation

Centers for Disease Control and Prevention. (2017, June 22). Rural and Urban Differences in Air Quality, 2008–2012, and Community Drinking Water Quality, 2010–2015 - United States. Centers for Disease Control and Prevention. https://www.cdc.gov/mmwr/volumes/66/ss/ss6613a1.htm. 

Centers for Disease Control and Prevention. (n.d.). National Environmental Public Health Tracking Network Query Tool. Centers for Disease Control and Prevention. https://ephtracking.cdc.gov/DataExplorer/. 

Hayes, A. (2020, November 11). What Is Cross-Correlation? Investopedia. https://www.investopedia.com/terms/c/crosscorrelation.asp. 

Prabhakaran, S. (n.d.). if(typeof __ez_fad_position != 'undefined'){__ez_fad_position('div-gpt-ad-r_statistics_co-box-3-0')};Time Series Analysis. Time Series Analysis With R. http://r-statistics.co/Time-Series-Analysis-With-R.html. 

R: Holt-Winters Filtering. (n.d.). https://stat.ethz.ch/R-manual/R-devel/library/stats/html/HoltWinters.html. 

robk@statmethods.net, R. K.-. (n.d.). Time Series and Forecasting. Quick-R: Time Series. https://www.statmethods.net/advstats/timeseries.html. 

What Is PM2.5 and Why Is It Important?20. airveda. (2017). https://www.airveda.com/blog/what-is-pm2-5-and-why-is-it-important#:~:text=Dust%2C%20spores%20and%20pollen%20are,take%20place%20in%20the%20atmosphere.


